---
title: "Inference Course Project - Part B"
author: "Joe Nguyen"
date: "22 Nov 2015"
output: pdf_document
---
<!-- Completed on 31 Oct 2015, but submitted for 22 Nov 2015 -->

In this project, we use simulation to explore inference and do some simple inferential data analysis. The project consists of two parts:

- A simulation exercise
- Basic inferential data analysis


## B. Inference on $ToothGrowth$ Data

Information about the dataset:

The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).


### 1. Load the $ToothGrowth$ data

Explore the dataset.
```{r, echo=TRUE}
data("ToothGrowth")
str(ToothGrowth)
rbind(head(ToothGrowth,2), tail(ToothGrowth,2))
```


### 2. Provide a Basic Summary of the Data

```{r, echo=TRUE, fig.height=4, fig.width=8}
summary(ToothGrowth)

library(ggplot2)
tg <- ggplot(ToothGrowth, aes(x = dose, y = len), col = supp) + 
    geom_point(shape = 1, size = 3) + 
    facet_wrap(~ supp)
tg
```

Figure 1. Response (len) based on dose (0.5, 1, 2)mg and supplement type (OJ, VC) from 10 guinea pigs.


### 3. Compare Tooth Growth

Tooth growth (len) is compared by dose and supplement type (supp). T-tests are used to determine the difference between pairs of dose and the supp pair. We use t-tests because the response (len) appears to by roughly symmetric and mound shaped in Fig. 1.


#### i. Dose

First, differences in len are evaluate for pairs of dose levels = {0.5, 1, 2} mg. The data are paired by dose and so three paired t-test are performed:

| t-test Index | Dose A | Dose B |
| ------------ | ------ | ------ |
| 1 | 0.5 | 1 |
| 2 | 0.5 | 2 |
| 3 | 1 | 2 |

Also, group variances are assumed to be unequal as they are unknown. Assuming unequal variance results in wider confidence intervals than equal variance, which is more conservative. The 95% confidence intervals (CIs) are:

```{r, echo=TRUE}
lenDose1 <- subset(ToothGrowth, dose %in% c(0.5,1))
lenDose2 <- subset(ToothGrowth, dose %in% c(0.5,2))
lenDose3 <- subset(ToothGrowth, dose %in% c(1,2))

rbind(
    t.test(len ~ dose, paired = TRUE, var.equal = FALSE, data = lenDose1)$conf,
    t.test(len ~ dose, paired = TRUE, var.equal = FALSE, data = lenDose2)$conf,
    t.test(len ~ dose, paired = TRUE, var.equal = FALSE, data = lenDose3)$conf
)
```

As the 95% CIs are entirely below zero, these t-tests suggest the smaller doses result in less tooth growth; or, increasing the dose increases tooth growth. Additionally, the p-values are small (<< 0.05), strongly suggesting more tooth growth with higher doses.

```{r, echo=TRUE}
rbind(
    t.test(len ~ dose, paired = TRUE, var.equal = FALSE, data = lenDose1)$p.value,
    t.test(len ~ dose, paired = TRUE, var.equal = FALSE, data = lenDose2)$p.value,
    t.test(len ~ dose, paired = TRUE, var.equal = FALSE, data = lenDose3)$p.value
)
```


#### ii. Supplement

Second, a paired t-test is performed to evaluate a difference in tooth growth (len) between the two supplments (supp = {OJ, VC}). Again, unequal variance is assumed to be conservative.

```{r, echo=TRUE}
t.test(len ~ supp, paired = TRUE, var.equal = FALSE, data = ToothGrowth)
```

The 95% CI = (1.409, 5.991) suggests the supplment OJ provides more tooth growth than supplment VC. This is corroborated by a small p-value = 0.00255.

<!--
t.test structure
----------------
t.test(Ha, H0, ...)

Intepretation
-------------
If CI does not contain zero, it is statistically significant that we can accept Ha over H0; otherwise, we fail to reject H0. Similarly, if p-value < 0.05 (or small), accept Ha over H0.

This structure holds for paired tests where the first entry can be considered the new hypothesis (Ha) for the test (new - baseline) (Ha - H0). So if CIs is negative (and p-value is large), the second entry pair is more statistically significant. 
-->

